Neural Networks for Neural Tumors

### 2025-06-02 latex/draft.tex
revised eval/conc

\section{\large Evaluation (1-2 pages)}
In this project, we used three learning models (CNN, SVM, and RF) and compared the results of each. The dataset comprises four categories: glioma, meningioma, pituitary, and no tumor.

@@

Diving into per-class specifics, we observe that for the label "glioma", CNN exhibited the highest precision (0.98) and recall (0.93), indicating strong identification capability with few false positives and a relatively low false negative rate. SVM matched CNN in precision (0.98) but had a lower recall (0.91), while RF also had a precision (0.98) but the lowest score in recall at 0.89. This suggests that both SVM and RF are more likely to miss "glioma" cases compared to CNN while matching its precision.

@@

In classifying "no tumor", RF achieved perfect precision (1.00), meaning that every time RF predicted "no tumor", the model was correct. However, its recall (0.93) was the lowest among the three, pointing to a higher rate of false negatives. CNN balanced both precision (0.97) and recall (0.98) well, outperforming both alternatives in overall reliability for this class.

For the label "pituitary", all three models performed strongly, with CNN leading in both precision and recall (0.99 each). SVM and RF followed closely with both metrics at 0.98 and 0.97-0.98, respectively, indicating that all models are highly effective in classifying "pituitary" tumors.

\section{\large Conclusion (1/2 page)}
The evaluation highlight CNN's strength in handling complex, high-dimensional image data, likely due to its hierarchical feature extraction capability. Its superior balance in both precision and recall across all tumor types supports its suitability for real-world clinical deployment where both false positives and false negatives carry significant risk.

SVM performed comparably well, especially in the detection of meningioma and pituitary tumors, and may offer a lighter-weight alternative in environments with limited computational resources. However, it showed reduced precision in some classes, making it slightly less robust overall.

RF, while achieving remarkable precision, particularly for the "no tumor" class, exhibited reduced recall across multiple classes. This suggests a tendency to make fewer false positives but at the expense of missing actual tumor cases, which is a very critical and life-threatening issue in medical diagnostics.

@@

Richard: Richard has completed an article for literature review. Over the coming weeks, he will work towards exploratory data analysis (EDA) and preparing the dataset for analysis through data pre-processing.

>>>>

\section{\large Evaluation (1-2 pages)}
In this project, three machine learning models—Convolutional Neural Networks (CNN), Support Vector Machines (SVM), and Random Forests (RF)—were trained and evaluated. In this section, the performance of these models are assessed using standard classification metrics, including accuracy, precision, recall, and F1 score. The dataset utilized for this evaluation consists of four classes: glioma, meningioma, pituitary tumor, and no tumor.

@@

Diving into per-class specifics, we observe that for the label "glioma", CNN exhibited the highest precision (0.98) and recall (0.93), indicating strong identification capability with few false positives and a relatively low false negative rate. SVM matched CNN in precision (0.98) but had a lower recall (0.91), while RF also had a precision (0.98) but the lowest recall score at 0.89. This suggests that both SVM and RF are more likely to miss "glioma" cases compared to CNN while matching its precision.

@@

In classifying "no tumor", RF achieved perfect precision (1.00), which means that every time RF predicted "no tumor", the model was correct. However, its recall (0.93) was the lowest among the three, pointing to a higher rate of false negatives. CNN balanced both precision (0.97) and recall (0.98) well, outperforming both alternatives in overall reliability for this class.

For the label 'pituitary', all three models performed strongly, with CNN leading both in precision and recall (0.99 each). SVM and RF followed closely with both metrics at 0.98 and 0.97-0.98, respectively, indicating that all models are highly effective in classifying "pituitary" tumors.

\section{\large Conclusion (1/2 page)}
The above evaluation highlights CNN's strength in handling complex, high-dimensional image data, likely due to its hierarchical feature extraction capability. Its superior balance in both precision and recall in all tumor types supports its suitability for real-world clinical deployment, where false positives and false negatives carry significant risk.

SVM performed comparably well, especially in the detection of meningioma and pituitary tumors, and may offer a lighter alternative in environments with limited computational resources. However, it showed reduced precision in some classes, which made it slightly less robust overall.

RF, while achieving remarkable precision, particularly for the "no tumor" class, exhibited reduced recall across multiple classes. This suggests a tendency to make fewer false positives, at the expense of missing actual tumor cases, which is a very serious and life-threatening issue in medical diagnostics.

@@

Richard: Richard contributed to the project by evaluating the performance of the three machine learning model using evaluation metrics such as accuracy, F1 score, precision, and recall. Based on these analyses, he provided the rationale for selecting the CNN model as the final model due to its superior performance relative to the alternatives. Additionally, he conducted a comprehensive literature review to summarize key developments in the field, which served as a foundational reference for the project. His contributions also included performing exploratory data analysis (EDA) and preparing the dataset for modeling through appropriate data preprocessing techniques.
